Variational Inference for Inverse Reinforcement Learning with Gaussian Processes

### 2018-11-21 mprop/mprop.tex
The literature review is in a readable state now

Variational inference has seen a recent increase in interest among academics,
with different approaches focusing on different goals: better time complexity,
handling a wider variety of models, making approximations more accurate, and
using more complex function approximation techniques (such as neural networks)
to infer local latent variable values without having to calculate them
individually for each data point \cite{DBLP:journals/corr/abs-1711-05597}. As
our IRL model is based on a GP, we will begin by reviewing some of the VI
approaches applied specifically to GP regression. Based on a recent review of
scalable GPs \cite{DBLP:journals/corr/abs-1711-05597}, we will concentrate on
\emph{stochastic variational sparse approximations}, as they have achieved
modelling accuracy close to that of the full GP with no approximations, with
many methods providing a time complexity of $\mathcal{O}(m^3)$. Below we provide
a short overview of various assumptions that have been used in approximating
\emph{sparse} GPs (i.e., GPs that use inducing points), following on a paper by
Qui\~{n}onero-Candela and Rasmussen \cite{DBLP:journals/jmlr/CandelaR05}.

@@

\begin{table}
  \centering
  \begin{tabular}{llllll}
    \toprule
    Authors, year & Inducing points & Hyperparameters & Complexity \\
    \midrule
    Titsias, 2009 \cite{DBLP:journals/jmlr/Titsias09} & variational & variational & $\mathcal{O}(nm^2)$ \\
    Hensman et al., 2013 \cite{DBLP:conf/uai/HensmanFL13} & fixed & variational & $\mathcal{O}(m^3)$ \\
    Gal et al., 2014 \cite{DBLP:conf/nips/GalWR14} & variational & variational & $\mathcal{O}(nm^2)$ \\
    Hoang et al., 2015 \cite{DBLP:conf/icml/HoangHL15} & fixed & fixed & $\mathcal{O}(m^3)$ \\
    Cheng and Boots, 2017 \cite{DBLP:conf/nips/ChengB17} & variational & variational & $\mathcal{O}(nm_\alpha + nm_\beta^2)$ \\
    Hensman et al., 2017 \cite{DBLP:journals/jmlr/HensmanDS17} & fixed & variational & $\mathcal{O}(nm)$ \\
    Peng et al., 2017 \cite{DBLP:conf/icml/PengZZQ17} & variational & variational & $\mathcal{O}(m^3)$ \\
    \bottomrule
  \end{tabular}
  \caption{Summary of relevant VI approximations to GPs. For both inducing
    points and hyperparameters, `fixed' means `chosen before the algorithm
    starts' and `variational' means `included amongst the variational
    parameters'. Hyperparameters $m_\alpha$ and $m_\beta$ refer to the number of
    bases used to represent the GP's mean and covariance, respectively.}
  \label{table:approximations}
\end{table}

% TODO: the real reason is that we want to use the prior
% TODO: the reasoning about inducing points is not clear either
Table \ref{table:approximations} further summarises some of the recent and/or
influential GP approximation approaches that might be suitable for our GP IRL
model. In order to derive a reliable ELBO, the inducing points should be either
fixed or modelled by a probability distribution (none of the approaches do
that), while hyperparameters cannot be variational (the reason will become clear
in Section \ref{sec:proposed_approach}). We would also like to avoid having the
hyperparameters fixed, as learning them efficiently would introduce a separate
problem. As the approach by Hoang et al. does not seem to be easily extendable
to modelled hyperparameters, it is unsuitable to our needs.

The variational Fourier features (VFF) algorithm by Hensman et al.
\cite{DBLP:journals/jmlr/HensmanDS17} is only defined for Mat\'ern kernels,
which is likely to be too restrictive for our situation (the original paper on
using GPs for IRL \cite{DBLP:conf/nips/LevinePK11} used the automatic relevance
detection kernel that has weights controlling how important each feature is).
While extending VFF to support a flexible class of kernels defined by Wilson and
Adams \cite{pmlr-v28-wilson13} is an interesting and promising avenue of work,
it is likely to be beyond the scope of this project.

The derivation of the ELBO in the work of Peng et al.
\cite{DBLP:conf/icml/PengZZQ17} relies primarily on the fact that the evidence
for a GP is a Gaussian, while in our case the evidence can be defined in several
ways depending on the model, but must always involve $\mathcal{D}$, making the
main idea of the paper inapplicable to IRL. Remaining papers can be rejected for
similar reasons, as they are all highly dependent on the specific situation of
approximating a GP regression. Thus we cannot simply apply a GP VI model as a
whole, although there might be some smaller ideas that we can use (some of which
will be mentioned in the next section).

Finally, since we expect $\posterior$ to be highly irregular, we would like our
approximation to be capable of representing a wide range of possible probability
distributions. The primary way of representing complex posteriors in VI is by
using \emph{normalising flows}, i.e., a collection of invertible
functions---parametrised by additional variational parameters---that are applied
to latent variables \cite{DBLP:conf/icml/RezendeM15}. Unfortunately, this
parametrisation also means that the gradient of the joint probability
distribution w.r.t. variational parameters $\nabla_{\bm\nu}\pfull$ is no longer
zero, making an analytic expression for the ELBO impossible using the usual
methods. While it may be possible to have an approximating distribution for the
flow parameters, it is uncertain how such an algorithm would behave, as it would
have to perform optimisation in a space with significantly more dimensions.

@@

@@

We would like to compare our approach to alternatives that support nonlinear
reward functions. According to a recent survey
\cite{DBLP:journals/corr/abs-1806-06877}, this leaves us with five algorithms:
\textsc{LEARCH} \cite{DBLP:journals/arobots/RatliffSB09}, \textsc{MmpBoost}
\cite{DBLP:conf/nips/RatliffBBC06}, the original GP IRL paper
\cite{DBLP:conf/nips/LevinePK11}, its deep GP extension
\cite{DBLP:conf/uai/JinDAS17}, and a neural network reward function
approximation \cite{wulfmeier2015maximum}. As the first two have already been
shown to underperform \cite{DBLP:conf/nips/LevinePK11}, we will focus on the
remaining three.

>>>>

Variational inference has seen a recent increase in interest among academics,
with different approaches focusing on different goals: better time complexity,
handling a wider variety of models, making approximations more accurate, and
using more complex function approximation techniques (such as NNs)
to infer local latent variable values without having to calculate them
individually for each data point \cite{DBLP:journals/corr/abs-1711-05597}. As
our IRL model is based on a GP, we will begin by reviewing some of the VI
approaches applied specifically to GP regression. Based on a recent review of
scalable GPs \cite{DBLP:journals/corr/abs-1711-05597}, we will concentrate on
\emph{stochastic variational sparse approximations}, as they have achieved
modelling accuracy close to that of the full GP with no approximations, with
many methods providing a time complexity of $\mathcal{O}(m^3)$. Below we provide
a short overview of various assumptions that have been used in approximating
\emph{sparse} GPs (i.e., GPs that use inducing points), following on a paper by
Qui\~{n}onero-Candela and Rasmussen \cite{DBLP:journals/jmlr/CandelaR05}.

@@

\begin{table}
  \centering
  \begin{tabular}{llllll}
    \toprule
    Authors, year & Inducing points & Hyperparameters & Complexity \\
    \midrule
    Titsias, 2009 \cite{DBLP:journals/jmlr/Titsias09} & variational & variational & $\mathcal{O}(nm^2)$ \\
    Hensman et al., 2013 \cite{DBLP:conf/uai/HensmanFL13} & fixed & variational & $\mathcal{O}(m^3)$ \\
    Gal et al., 2014 \cite{DBLP:conf/nips/GalWR14} & variational & variational & $\mathcal{O}(nm^2)$ \\
    %Hoang et al., 2015 \cite{DBLP:conf/icml/HoangHL15} & fixed & fixed & $\mathcal{O}(m^3)$ \\
    Cheng and Boots, 2017 \cite{DBLP:conf/nips/ChengB17} & variational & variational & $\mathcal{O}(nm_\alpha + nm_\beta^2)$ \\
    Hensman et al., 2017 \cite{DBLP:journals/jmlr/HensmanDS17} & fixed & variational & $\mathcal{O}(nm)$ \\
    Peng et al., 2017 \cite{DBLP:conf/icml/PengZZQ17} & variational & variational & $\mathcal{O}(m^3)$ \\
    \bottomrule
  \end{tabular}
  \caption{Summary of relevant VI approximations to GPs. For both inducing
    points and hyperparameters, `fixed' means `chosen before the algorithm
    starts' and `variational' means `included amongst the variational
    parameters'. Hyperparameters $m_\alpha$ and $m_\beta$ refer to the number of
    bases used to represent the GP's mean and covariance, respectively.}
  \label{table:approximations}
\end{table}

Table \ref{table:approximations} further summarises some of the recent and/or
influential GP approximation approaches that might be relevant to our situation.
For many of the variables in our model (including the inducing points
located somewhere in $\mathbb{R}^d$ and the hyperparameters of the covariance
function), we have three possible ways of handling them:
\begin{itemize}
\item Provide a full Bayesian treatment by defining a prior probability
  distribution. While this option can easily prevent overfitting, the choice of
  prior can be difficult to justify.
\item Fix the value of the variable. We can then optimise the value outside of
  the main algorithm, e.g., by running the algorithm with a set of possible
  values and keeping the value that produces the best performance. This option
  can be slow if we are interested in finding the optimal value, but could be
  appropriate if the variable has little impact on the algorithm's performance.
\item Treat the variable as a variational parameter. This method will
  efficiently optimise the value of the variable, but is vulnerable to
  overfitting. Furthermore, this requires us to find the derivative of the ELBO
  w.r.t. the variable, which could be difficult.
\end{itemize}

Titsias \cite{DBLP:journals/jmlr/Titsias09} is the first to suggest a
variational approximation where inducing points are treated as variational
parameters. The paper has influenced many of the later works covered in this
section, and has been reformulated by Gal et al. \cite{DBLP:conf/nips/GalWR14}
into a distributed algorithm. We will fix the inducing points for the baseline
version of our algorithm, but keep the idea in mind for possible future work.

Hensman et al. \cite{DBLP:conf/uai/HensmanFL13} reduce the complexity from
$\mathcal{O}(nm^2)$ to $\mathcal{O}(m^3)$ using a stochastic variational
inference (SVI) approach. While results such as efficiently computable natural
gradients and analytically tractable optimal solutions for variational
parameters are unlikely to transfer to a different problem domain, the overall
SVI framework and an approximating Gaussian distribution for $\mathbf{u}$ will
play an important role in our proposal.

Cheng and Boots \cite{DBLP:conf/nips/ChengB17} suggest using different bases in
the reproducing kernel Hilbert space for the mean and covariance functions.
The paper makes few assumptions and mostly relies on the work of
Titsias \cite{DBLP:journals/jmlr/Titsias09}, making it seem transferable to a
new domain. However, applying the idea to IRL is likely to be beyond the scope
of the project.

The variational Fourier features (VFF) algorithm by Hensman et al.
\cite{DBLP:journals/jmlr/HensmanDS17} is only defined for Mat\'ern kernels,
which is likely to be too restrictive for our situation (the original paper on
using GPs for IRL \cite{DBLP:conf/nips/LevinePK11} used the automatic relevance
detection kernel that has weights controlling how important each feature is).
While extending VFF to support a flexible class of kernels defined by Wilson and
Adams \cite{pmlr-v28-wilson13} is an interesting and promising avenue of work,
it is likely to be beyond the scope of this project as well.

While Peng et al. \cite{DBLP:conf/icml/PengZZQ17} provide a highly efficient
distributed implementation, the derivation of the ELBO relies primarily on the
fact that the evidence for a GP is a Gaussian, while in our case the evidence is
anything but. Hence, it is unlikely that the ideas from this paper could be
applicable to our IRL model.

Finally, since we expect $\posterior$ to be highly irregular, we would like our
approximation to be capable of representing a wide range of possible probability
distributions. The primary way of representing complex posteriors in VI is by
using \emph{normalising flows}, i.e., a collection of invertible
functions---parametrised by additional variational parameters---that are applied
to latent variables \cite{DBLP:conf/icml/RezendeM15}. Normalising flows, along
with some of the ideas mentioned previously, could provide great benefits to the
GP IRL VI model. However, we are forced to keep any non-essential features as
possible future work in order to make the project feasible in a given time
frame. Furthermore, note that due to various differences between regression and
IRL, we cannot simply apply a GP VI model as a whole. Instead, smaller tricks
and ideas from various papers can (and will) be used throughout the project.

@@

% TODO: the real reason is that we want to use the prior

@@

We would like to compare our approach to alternatives that support nonlinear
reward functions. According to a recent survey
\cite{DBLP:journals/corr/abs-1806-06877}, this leaves us with five algorithms:
\textsc{LEARCH} \cite{DBLP:journals/arobots/RatliffSB09}, \textsc{MmpBoost}
\cite{DBLP:conf/nips/RatliffBBC06}, the original GP IRL paper
\cite{DBLP:conf/nips/LevinePK11}, its deep GP extension
\cite{DBLP:conf/uai/JinDAS17}, and an NN-based reward function
approximation \cite{wulfmeier2015maximum}. As the first two have already been
shown to underperform \cite{DBLP:conf/nips/LevinePK11}, we will focus on the
remaining three.
